\sys: Automatic Disease Diagnosis Using Large Language Models and Answer Set Programming

### 2025-06-14 paper/draft.tex
Improvements across
Signed-off-by: Evangelos Lamprou <vagos@lamprou.xyz>

\usepackage{datetime}
\usepackage{wrapfig}
\usepackage{enumerate}
\usepackage{graphicx}
\usepackage{caption}
\usepackage{siunitx}
\usepackage{subcaption}
\usepackage{amsmath}
\usepackage{array,tabularx}
\usepackage{chngcntr}
\usepackage{afterpage}
\usepackage{ulem}
\usepackage{hyperref}
\usepackage{dirtytalk}
\usepackage{algorithm2e}
\usepackage{enumitem,amssymb}
\usepackage{pifont}
\usepackage{amsmath}
\usepackage{listings}
\usepackage{xcolor}
\usepackage{formal-grammar}
\usepackage{varwidth}
\usepackage{fmtcount}
\usepackage{tikz}
\usepackage{array}
\usepackage[framemethod=tikz]{mdframed} % Allows defining custom boxed/framed environments
\usepackage{cleveref}
\usepackage{microtype}
\usepackage{dirtytalk}

@@

\hypersetup{
    colorlinks=true,
    filecolor=magenta,      
    urlcolor=blue,
}

@@

\title{Automatic Disease Diagnosis Using Large Language Models and Answer Set Programming}
\author{
	Ioanna Gemou
	\and
	Evangelos Lamprou
}

@@

\maketitle
\begin{abstract}
Accurate disease prediction is vital for timely intervention, 
effective treatment, and reducing medical complications. 
While symbolic AI has been applied in healthcare, 
its adoption remains limited due to the complexity 
of required knowledge bases.
This work introduces \sys, a framework that combines Large Language Models (LLMs) 
with Answer Set Programming (ASP) to overcome this barrier. 
LLMs extract medical knowledge from text and translate it into ASP code, 
while ASP enables structured, explainable reasoning. 
This integration yields a robust, interpretable prediction framework 
that leverages the strengths of both paradigms. 
\sys shows strong performance on real patient data.
% \sys accurately predicts diseases based on symptoms derived from medical literature.
\end{abstract}

% TODO: Add text between sections and paragraphs
\section{Introduction}

@@

This paper introduces \sys, a framework 
that combines Large Language Models (LLMs) with the 
formal reasoning capabilities of ASP.
\sys extracts and encodes medical knowledge from text into ASP code,
enabling interpretable and accurate diagnosis 
of diseases based on patient symptoms.

@@

\paragraph{\textbf{Large Language Models}}
Large language models (LLMs) \cite{zhao2023survey} are advanced computational systems 
designed to understand and generate human language. 
Developed using cutting-edge machine learning techniques, 
these models can produce text that closely mirrors human linguistic patterns.
Modern architectures such as \textit{GPT-4} \cite{openai2023gpt4}, 
are trained on massive corpora spanning diverse domains, 
enabling them to acquire broad, cross-disciplinary knowledge. 
LLMs use deep learning algorithms \cite{Sarker2021} 
and specifically transformer networks \cite{Dosovitskiy2020} 
to analyze and understand the relationships between words, phrases, and sentences. 
By capturing patterns and structures in the training data, 
they can generate coherent and relevant responses based on given prompts.
The term “large” refers to the billions of parameters 
that allow these models to generate context-aware responses. 
The basic premise of a language model is its ability to predict 
the next word (referred to as a token) based on the text it has observed so far, 
according to the context.

@@

\begin{figure}[htb]
    \begin{center}
\begin{tikzpicture}[node distance=4cm]
        \node [simple-rect] (problem)  at (0,0) {Problem};
        \node [simple-rect, below of=problem] (program) {Program};

        \node [simple-rect, right of=program] (grounder) {Grounder};
        \node [simple-rect, right of=grounder] (solver) {Solver};

@@

        \node[draw, dotted, fit=(grounder) (solver), inner sep=4mm, label=above:{Solving}] {};
\end{tikzpicture}
    \end{center}
		\caption{Solving a problem using ASP.}
    \label{fig:asp-solving}
\end{figure}

@@

\begin{itemize}
    \item \textbf{Rule-Based Representation}: ASP allows for the representation of medical 
    knowledge as rules in the form of logical statements. 
    These rules define relationships between medical entities such as symptoms, 
    diseases, treatments, and patient data. For example, a rule could state that 
    if a patient exhibits specific symptoms, it indicates the presence or absence 
    of certain diseases.
    \item \textbf{Knowledge Base Construction}: ASP facilitates the construction of 
    structured knowledge bases composed of facts, rules, and constraints. These can 
    model diagnostic criteria, symptom-disease associations, and therapeutic protocols 
    in a consistent and extensible manner.
\end{itemize}

@@

Recent work explores the synergy between LLMs and ASP 
for enhanced diagnostic reasoning. 
These hybrid approaches aim to combine the language understanding of LLMs 
with the precision and explainability of symbolic reasoning. 
A YAML-based interface has been proposed to guide LLMs 
in extracting structured facts from natural language, 
which are then processed by ASP for logical inference \cite{alviano2024llm2asp}. 
Other work uses ASP to validate and trace misleading or inconsistent outputs 
from LLMs in medical contexts \cite{Nguyen2025}. 
A specialized lightweight LLM has also been trained to generate valid ASP code, 
improving the automation of rule-based medical reasoning \cite{coppolillo2024llasp}.
These systems assist in building reliable and clear reasoning pipelines 
for disease diagnosis by combining language understanding with rule-based validation.
\section{Methodology}

@@

The tool \textit{xclingo}\footnote{\url{https://github.com/bramucas/xclingo2}}~\cite{Cabalar_2020} 
enables the visualization of justifications for results produced by the \textit{Clingo} solver. 
It does so by incorporating a \src{trace} that corresponds to the rules of the logic program. 
Providing clear explanations of how the framework reaches specific conclusions helps users identify 
potential errors in the construction of the knowledge base and offers transparency for the final outcome. 
This is especially important in medical contexts, where interpretability and trust are critical.

@@

\begin{table}[h]
    \centering
    \begin{tabular}{lcc}
    \hline
    Diseases & Size of Knowledge Base & Accuracy \\
    \hline
    \hline
    Chicken Pox    &  66  & 95\% \\
    Pneumonia      &  75  & 100\% \\
    Common Cold    &  44  & 100\% \\
    \hline
    \end{tabular}
    \caption{Accuracy of the framework on selected diseases. 
    The size column indicates the number of logical terms in the disease-specific ASP program.}
\end{table}

@@

\begin{lstlisting}[caption={Excerpt from the explanation tree for the chickenpox diagnosis. The structure illustrates the reasoning path based on symptom associations (slightly modified for clarity).},
    label=lst:chickenpox_explanation]{Name}
    *
    |__ diagnosis(chickenpox)
        |__ has(symptom(itching))
        |__ has(symptom(fatigue))
        |__ has(symptom(lethargy))
            |__ has(symptom(fatigue))
            |__ linked_symptom(fatigue, lethargy)
        |__ has(symptom(high_fever))
            |__ has(symptom(mild_fever))
                |__ has(symptom(loss_of_appetite))
                |__ linked_symptom(loss_of_appetite, mild_fever)
            |__ linked_symptom(mild_fever, high_fever)
        |__ has(symptom(loss_of_appetite))
        |__ has(symptom(mild_fever))
            |__ has(symptom(loss_of_appetite))
            |__ linked_symptom(loss_of_appetite, mild_fever)
        |__ has(symptom(swelled_lymph_nodes))
\end{lstlisting}

\Cref{lst:chickenpox_explanation} shows an excerpt from the explanation 
tree generated for a patient diagnosed with \textit{chickenpox}. 
The tree visualizes the reasoning steps performed by the ASP solver to reach this conclusion. 
At the root, \sys infers \texttt{diagnosis(chickenpox)} 
based on the presence of multiple symptoms such as \texttt{itching}, \texttt{fatigue}, \texttt{lethargy}, 
and \texttt{high\_fever}. 
Each symptom is further justified through observed inputs or inferred relationships. 
For example, \texttt{lethargy} is supported by repeated instances of \texttt{fatigue} 
and the known association \texttt{linked\_symptom(fatigue, lethargy)}. 
Similarly, \texttt{high\_fever} is explained via the intermediate symptom \texttt{mild\_fever}, 
which itself is supported by \texttt{loss\_of\_appetite}. 
These explanations demonstrate how \sys combines direct observations 
and transitive symptom relationships encoded in the knowledge base to 
construct a transparent and interpretable diagnostic path.

>>>>

\usepackage{datetime}
\usepackage{wrapfig}
\usepackage{enumerate}
\usepackage{graphicx}
\usepackage{caption}
\usepackage{siunitx}
\usepackage{subcaption}
\usepackage{amsmath}
\usepackage{array,tabularx}
\usepackage{chngcntr}
\usepackage{afterpage}
\usepackage{ulem}
\usepackage{hyperref}
\usepackage{dirtytalk}
\usepackage{algorithm2e}
\usepackage{enumitem,amssymb}
\usepackage{pifont}
\usepackage{amsmath}
\usepackage{listings}
\usepackage{xcolor}
\usepackage{formal-grammar}
\usepackage{varwidth}
\usepackage{fmtcount}
\usepackage{tikz}
\usepackage{array}
\usepackage[framemethod=tikz]{mdframed} % Allows defining custom boxed/framed environments
\usepackage{cleveref}
\usepackage{microtype}
\usepackage{dirtytalk}
\usepackage{booktabs}

@@

\hypersetup{
    colorlinks=true,
    filecolor=magenta,      
		citecolor=black,
    urlcolor=blue,
}

@@

\makeatletter
\def\@settitle{%
  \begin{center}%
    \normalfont\LARGE\bfseries
    \@title\par
  \end{center}%
}
\makeatother

\title[\sys: Automatic Disease Diagnosis Using LLMs \& ASP]{\sys: Automatic Disease Diagnosis Using Large Language Models and Answer Set Programming}

\author[Ioanna Gemou]{Ioanna Gemou}
\author[Evangelos Lamprou]{Evangelos Lamprou}
\thanks{Originally submitted as part of a research competition on 1st June 2023.}

@@

\maketitle
\vspace{-1.5em} % adjust as needed
{\small\itshape \centering Originally submitted as part of a university research competition on 1st June 2023.\par}
\vspace{0.5em} % optional: add space before body

\begin{abstract}
Accurate disease prediction is vital for timely intervention, 
effective treatment, and reducing medical complications. 
While symbolic AI has been applied in healthcare, 
its adoption remains limited due to the complexity 
of required knowledge bases.
This work introduces \sys, a framework that combines Large Language Models (LLMs) 
with Answer Set Programming (ASP) to overcome this barrier. 
LLMs extract medical knowledge from text and translate it into ASP code, 
while ASP enables structured, explainable reasoning. 
This integration yields a robust, interpretable prediction framework 
that leverages the strengths of both paradigms. 
\sys shows strong performance on real patient data.
% \sys accurately predicts diseases based on symptoms derived from medical literature.
\end{abstract}

\section{Introduction}

@@

This paper introduces \sys, a framework 
that combines LLMs with the 
formal reasoning capabilities of ASP.
\sys extracts and encodes medical knowledge from text into ASP code,
enabling interpretable and accurate diagnosis 
of diseases based on patient symptoms.

@@

\paragraph{\textbf{Large Language Models}}
LLMs \cite{zhao2023survey} are advanced computational systems 
designed to understand and generate human language. 
Developed using cutting-edge machine learning techniques, 
these models can produce text that closely mirrors human linguistic patterns.
Modern architectures such as \textit{GPT-4} \cite{openai2023gpt4}, 
are trained on massive corpora spanning diverse domains, 
enabling them to acquire broad, cross-disciplinary knowledge. 
LLMs use deep learning algorithms \cite{Sarker2021} 
and specifically transformer networks \cite{Dosovitskiy2020} 
to analyze and understand the relationships between words, phrases, and sentences. 
By capturing patterns and structures in the training data, 
they can generate coherent and relevant responses based on given prompts.
The term “large” refers to the billions of parameters 
that allow these models to generate context-aware responses. 
The basic premise of a language model is its ability to predict 
the next word (referred to as a token) based on the text it has observed so far, 
according to the context.

@@

\begin{figure}[htb]
    \begin{center}
\begin{tikzpicture}[node distance=2.9cm]
        \node [simple-rect] (problem)  at (0,0) {Problem};
        \node [simple-rect, below of=problem] (program) {Program};

        \node [simple-rect, right of=program] (grounder) {Grounder};
        \node [simple-rect, right=2cm of grounder] (solver) {Solver};

@@

        \node[draw, dotted, fit=(grounder) (solver), inner sep=2mm, label=above:{Solving}] {};
\end{tikzpicture}
    \end{center}
		\caption{Solving a problem using ASP.}
    \label{fig:asp-solving}
\end{figure}

@@

\begin{itemize}
    \item \textbf{Rule-Based Representation}: ASP allows for the representation of medical 
    knowledge as rules in the form of logical statements. 
    These rules define relationships between medical entities such as symptoms, 
    diseases, treatments, and patient data. For example, a rule could state that 
    if a patient exhibits specific symptoms, it indicates the presence or absence 
    of certain diseases.
		There have been multiple instances of rule-based medical reasoning systems~\cite{CHEN_2016,marple2017s}
    \item \textbf{Knowledge Base Construction}: ASP facilitates the construction of 
    structured knowledge bases composed of facts, rules, and constraints. These can 
    model diagnostic criteria, symptom-disease associations, and therapeutic protocols 
		in a consistent and extensible manner~\cite{chen2018automating,CHEN_2016}.
\end{itemize}
\sys builds upon these works and identifies a key bottleneck, which is the construction of a
knowledge base that can cover a wide range of diseases and conditions.

@@

More recent work explores the synergy between LLMs and ASP 
for enhanced diagnostic reasoning. 
These hybrid approaches aim to combine the language understanding of LLMs 
with the precision and explainability of symbolic reasoning. 
A YAML-based interface has been proposed to guide LLMs 
in extracting structured facts from natural language, 
which are then processed by ASP for logical inference \cite{alviano2024llm2asp}. 
Other work uses ASP to validate and trace misleading or inconsistent outputs 
from LLMs in medical contexts \cite{Nguyen2025}. 
A specialized lightweight LLM has also been trained to generate valid ASP code, 
improving the automation of rule-based medical reasoning \cite{coppolillo2024llasp}.
These systems assist in building reliable and clear reasoning pipelines 
for disease diagnosis by combining language understanding with rule-based validation.
These systems were developed after \sys and illustrate the convergent development of similar approaches, highlighting the intuitive appeal and practical value of the underlying techniques.

\section{Methodology}

@@

The tool \textit{xclingo}~\cite{Cabalar_2020} 
enables the visualization of justifications for results produced by the \textit{Clingo} solver. 
It does so by incorporating a \src{trace} that corresponds to the rules of the logic program. 
Providing clear explanations of how the framework reaches specific conclusions helps users identify 
potential errors in the construction of the knowledge base and offers transparency for the final outcome. 
This is especially important in medical contexts, where interpretability and trust are critical.

@@

\begin{table}[h]
    \centering
    \caption{Accuracy of the framework on selected diseases. 
    The size column indicates the number of logical terms in the disease-specific ASP program.}
    \begin{tabular}{lcc}
    \toprule
    Diseases & Size & Accuracy \\
    \midrule
    Chicken Pox    &  66  & 95\% \\
    Pneumonia      &  75  & 100\% \\
    Common Cold    &  44  & 100\% \\
    \bottomrule
    \end{tabular}
\end{table}

@@

\begin{lstlisting}[caption={Excerpt from the explanation tree for the chickenpox diagnosis. The structure illustrates the reasoning path based on symptom associations (slightly modified for clarity).},
    label=lst:chickenpox-explanation]{Name}
    *
    |__ diagnosis(chickenpox)
        |__ has(symptom(itching))
        |__ has(symptom(fatigue))
        |__ has(symptom(lethargy))
            |__ has(symptom(fatigue))
            |__ linked_symptom(fatigue, lethargy)
        |__ has(symptom(high_fever))
            |__ has(symptom(mild_fever))
                |__ has(symptom(loss_of_appetite))
                |__ linked_symptom(loss_of_appetite, mild_fever)
            |__ linked_symptom(mild_fever, high_fever)
        |__ has(symptom(loss_of_appetite))
        |__ has(symptom(mild_fever))
            |__ has(symptom(loss_of_appetite))
            |__ linked_symptom(loss_of_appetite, mild_fever)
        |__ has(symptom(swelled_lymph_nodes))
\end{lstlisting}

\Cref{lst:chickenpox-explanation} shows an excerpt from the explanation 
tree generated for a patient diagnosed with \textit{chickenpox}. 
The tree visualizes the reasoning steps performed by the ASP solver to reach this conclusion. 
At the root, \sys infers \texttt{diagnosis(chickenpox)} 
based on the presence of multiple symptoms such as \texttt{itching}, \texttt{fatigue}, \texttt{lethargy}, 
and \texttt{high\_fever}. 
Each symptom is further justified through observed inputs or inferred relationships. 
For example, \texttt{lethargy} is supported by repeated instances of \texttt{fatigue} 
and the known association \texttt{linked\_symptom(fatigue, lethargy)}. 
Similarly, \texttt{high\_fever} is explained via the intermediate symptom \texttt{mild\_fever}, 
which itself is supported by \texttt{loss\_of\_appetite}. 
These explanations demonstrate how \sys combines direct observations 
and transitive symptom relationships encoded in the knowledge base to 
construct a transparent and interpretable diagnostic path.
